---
title: "SC1102: R Bootcamp Week 2"
output:
  html_document:
    toc: true
    toc_depth: '4'
  pdf_document:
    toc: true
    toc_depth: '4'
date: "2026 Trimester 3"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[&larr; Back to the SC1102 course page](../)

> **DRAFT OUTLINE — not yet written.** This page is a structural skeleton for Week 2.
> Section headings, learning objectives and the intended code for each step are laid out
> below so the shape of the session can be reviewed before the prose is written.

## Writing scripts and reproducibility ##

**Goal:** students stop typing into the console and start writing scripts that they, and anyone else, can re-run from scratch and get an identical answer.

**The one-sentence version:** last week you made R do things; this week you make R do things *again*.

---

## Lecture (1 hour) — Why anyone should be able to re-run your work ##

*Live-coded, as per Week 1.*

### Learning objectives ###

- Explain what "reproducible" means for a data analysis, and why science depends on it.
- Describe the difference between the console and a script, and when to use each.
- Explain what a working directory is and why absolute file paths break other people's computers.
- Set up an RStudio Project with a sensible folder structure.
- Recognise the habits that quietly make an analysis irreproducible.

### Topics to cover ###

**1. The console is a whiteboard; the script is the record.**
Everything from Week 1 is gone the moment you close RStudio. Nothing you did is recoverable, checkable or shareable. That is fine for exploring and fatal for science.

**2. What reproducibility actually means.**
Same data + same code → same result, obtained by someone else, on a different machine, later. Three failure modes to name explicitly:

- I can't re-run it (the steps only existed in my head or in my console history)
- You can't re-run it (it depends on something only on my machine)
- Nobody can re-run it (the raw data got edited in place and the original is gone)

**3. Cautionary tale.**
The Reinhart–Rogoff spreadsheet error (Herndon, Ash & Pollin, 2013) — an averaging range in Excel that omitted several countries, in a paper used to argue for austerity policy. The error was only found because someone eventually got hold of the actual spreadsheet.

> **VERIFY BEFORE LECTURING** — this story is flagged in the project notes as needing a source check. Herndon, Ash & Pollin (2013), *Cambridge Journal of Economics*, is the critique paper. Confirm the specific details you plan to state out loud.

**4. The working directory — R's "you are here".**

```{r eval=FALSE}
getwd()      # where am I?
list.files() # what can I see from here?
```

Absolute vs relative paths. Why this line is a bug, not a path:

```{r eval=FALSE}
# Don't do this. It works on exactly one computer in the world.
fisheries <- read.csv("/home/andrew/Documents/teaching/2026/TR3/SC1102/fisheries.csv")
```

> **Teaching note — use the real example.** The first version of the Week 1 tutorial had exactly this
> line in it, and it broke the moment the file moved to a different machine. Worth showing: this is a
> mistake everyone makes, including the person writing the tutorial.

**5. RStudio Projects as the fix.**
A `.Rproj` file marks the top of your project. Open the project and the working directory is set for you, every time, on every machine. Relative paths then just work.

**6. A folder structure that scales.**

```
SC1102/
├── SC1102.Rproj
├── data/          <- raw data. Read from here. NEVER write to here.
├── scripts/       <- your .R files
├── figures/       <- plots your script produces
└── outputs/       <- processed data your script produces
```

The rule worth repeating all week: **raw data is read-only.** If you need to fix something, fix it in code, not in the file.

**7. Habits that quietly break reproducibility.**

- Saving and restoring `.RData` (turn it off: Tools > Global Options > uncheck "Restore .RData", set "Save workspace" to Never)
- Relying on objects that are in your Environment but not in your script
- `rm(list = ls())` — feels like a reset, isn't one (loaded packages and options survive). Session > Restart R is the real thing.
- Running lines out of order until it works

**8. `sessionInfo()`** — recording what versions of everything you used.

> **Excel bridge.** In Excel, the analysis and the data are the same object, so "re-running" it isn't a thing you can even ask for. In R the data sits still and the script is the thing you keep, edit, share and re-run. That separation is the entire point.

---

## Practical (2 hours) — Build a project, write a script ##

Students convert last week's console session into a real script inside a real project. Pair programming; predict-then-run.

### Step 1 — Create the Project ###

File > New Project > New Directory > New Project. Name it `SC1102`. Then create the subfolders and move `fisheries.csv` into `data/`.

```{r eval=FALSE}
dir.create("data")
dir.create("scripts")
dir.create("figures")
dir.create("outputs")
```

### Step 2 — Start a script with a header block ###

```{r eval=FALSE}
# ------------------------------------------------------------
# Week 2: cleaning the fisheries catch data
# Author:  <your name>
# Date:    2026-##-##
# Purpose: read the raw catch records, fix the column types,
#          and save a cleaned version for later analysis.
# ------------------------------------------------------------
```

### Step 3 — Read with a relative path ###

```{r eval=FALSE}
fisheries <- read.csv("data/fisheries.csv")   # 2000 catch records
str(fisheries)
```

### Step 4 — Rebuild Week 1's work as a script ###

Type the type-conversions from Week 1 into the script rather than the console. Same code, permanent home.

```{r eval=FALSE}
fisheries$date       <- as.Date(fisheries$date)
fisheries$legal_size <- as.logical(fisheries$legal_size)
```

### Step 5 — The real test: restart and source ###

Session > Restart R (Ctrl/Cmd + Shift + F10), then run the whole script from a clean slate.

```{r eval=FALSE}
source("scripts/week_2.R")
```

If it runs, it is reproducible. If it doesn't, you have just found a dependency you didn't know you had. This step is the point of the whole session.

### Step 6 — Break it and fix it ###

Deliberately sabotage the script and read the error:

| Sabotage | Error students should learn to read |
|---|---|
| Use an absolute path | `cannot open file ... No such file or directory` |
| Misspell the filename | same error — teaches that R is not guessing for you |
| Move a line above the one that creates its input | `object 'fisheries' not found` |
| Delete a closing bracket | `unexpected symbol` / console stuck on `+` |

### Step 7 — Reading the Excel file directly ###

```{r eval=FALSE}
library(readxl)
fisheries_xl <- read_excel("data/fisheries.xlsx")
```

Compare with `read.csv()`. Note that `read_excel()` guesses column types differently — another reason to always run `str()` after reading anything in.

### Step 8 — Building text with `paste()` ###

```{r eval=FALSE}
# A label for a figure or a site code
paste(fisheries$site, fisheries$vessel, sep = "_")

# A filename that records when it was made
outfile <- paste0("outputs/fisheries_clean_", Sys.Date(), ".csv")
```

### Step 9 — Write outputs from code, never by hand ###

```{r eval=FALSE}
write.csv(fisheries, outfile, row.names = FALSE)

png("figures/undersized_by_vessel.png", width = 800, height = 600)
barplot(prop_tab[, "FALSE"],
        ylab = "Proportion of undersized catches",
        xlab = "Vessel ID",
        ylim = c(0, 0.5))
dev.off()
```

The point: the figure in your report is produced by the script, so it can never drift out of date.

### Step 10 — Don't repeat yourself: writing a function ###

> **OPEN DECISION — see COVERAGE.md.** Functions have to live somewhere in the new three-week
> structure. They sit naturally here ("you have written the same three lines three times — wrap them
> up"), but Week 2 is already a full session. The alternative is to move functions into Week 3 as part
> of building the report. Flagged for Andrew to choose.

```{r eval=FALSE}
# What a function is: a named recipe that takes inputs and returns an output.
# You have already used dozens: mean(), read.csv(), as.Date().

# Defining one
kg_to_lb <- function(kg) {
  kg * 2.205        # mass in lb
}

# Executing it
kg_to_lb(3.71)
fisheries$weight_lb <- kg_to_lb(fisheries$weight_kg)
```

### Step 11 — Record your session ###

```{r eval=FALSE}
sessionInfo()
```

> **End-of-week win.** Finish by deleting the `figures/` folder, restarting R, sourcing the script,
> and watching the figure rebuild itself from nothing. That moment — the analysis regenerating
> itself — is what reproducibility feels like.

---

## Synthesis (1 hour) — Can someone else run your work? ##

- **The swap.** Pairs exchange project folders. Can you run your partner's script without editing a single line? Anything you had to change is a reproducibility bug — name it and fix it together.
- **Failure gallery.** Work through the classic breakages as a class: absolute paths, missing `data/` folder, an object that only existed in the Environment, a package that was never loaded.
- **Discussion.** If a journal asked for your analysis in five years, what would you need to hand over? What would still work?
- **Exit ticket.** One habit from this week you will keep, and one thing still confusing.
- **Preview Week 3.** *"Right now your analysis and your write-up are two separate things, and the numbers in your write-up are already going stale. Next week they become one document."*

---

*SC1102 R Bootcamp, James Cook University. [Course home](../)*
